An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
wrote about cost vs. random initialization

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but with randomly permutated costs where each occuring cost value has been divided by a random value between 0 and 5000.

@@

@@

\subsection{Random Initialization Versus Cost-Optimal Initialization}
In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above. Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations. Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself. \TODO{However, what can be seen is that the ...}

Boxplots can be found in Fig.~\ref{fig:costVsRandBoxPlots} ...

@@

\subsection{Changing the Population Size}
In a second experiment, we show that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically. To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities. From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favour. When comparing the hypervolume values of the resulting populations, however, an effect is visible. Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveil statistical significant differences between two of the three possible algorithm pairs. With a $p$-value of $0.02163$, the null hypothesis of no differences between a population size of 50 and a population size of 100 has to be rejected in favor of stating that the algorithm variant with population size 50 is superior to the one with population size 100. For the comparison between population size 50 and population size 200, the $p$-value equals $0.0144$ with the population size 50 being better whereas the comparison between population size 100 and population size 200 does not reveal a statistical difference (the $p$-values of the one-sided Wilcoxon tests are $0.2179$ and $0.8035$ respectively).

@@

\begin{figure}
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}%
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}%
	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{S_FW}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest. Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary algorithm to cope with the non-linear equity risk objective.

\TODO{ In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.
}

The third objective function of our problem can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$. Although this equivalent formulation makes the problem linear (with additional linear constraints), classical algorithms are expected to have difficulties with this formulation as well and our algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints. Note that, for the moment, the proposed EMO algorithm exists on paper only and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to generate solutions of sufficient quality and whether adaptively changing the number and capacity of trucks is beneficial.

In future work, one might want to look further into adding (simple) heuristics within the evoluionary algorithm in order to improve the algorithm. For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cylces. With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.

% The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest.

% In this study, we formalize transportation of hazmats problem as the minimization of three objectives and propose to use an evolutionary algorithm since many exact optimization procedures have difficulties with the non-linear equity risk objective. Note that our problem formulation can be rewritten by minimizing the additional variable $z$ as third objective and adding the constraints $\forall q \in Q: z \geq \sum_{c \in C} \sum_{(i,j) \in A} r_{ij}^{cq} f_{ij}^c$ \COMMENTD{this has to be adapted according to the new problem formulation}. Although this equivalent formulation makes the problem linear (with additional non-linear constraints), classical algorithms are expected to have difficulties with this formulation as well and the EMO algorithm is supposed to be more efficient in the current formulation due to the fewer number of constraints.
% 
% Note also that, for the moment, the proposed EMO algorithm only exists as a gedankenexperiment and an actual implementation has to prove in the future which additional algorithm components (such as problem-specific initialization, recombination operators, or other exact optimization (sub-)procedures) are necessary to improve the quality of the found solutions to a sufficient level.

>>>>

As a test case for all the experiments, we chose to use a real-world case study based on the work of \citet{bcg2009a}. The network used here consists of 311 nodes and 441 arcs and represents the main transport roads in Italy's Lazio region which, amongst other cities, includes the capital Rome. We distinguish between instances with a different number of commodities (between 2 and 4) and with different numbers of trucks, see Table~\ref{tab:instances}. While the costs $c_{ij}$ are proportional to the real distance between the two locations $i$ and $j$, the risks $r_{ij}^c$ for each arc $(i,j) \in A$ and commodity $c \in C$ have been computed as a product of the arc's length, the actual population density in the area of the arc and a constant which depends on the commodity, for details see \citep{bcg2009a} and Table~\ref{tab:instances}. Solely one instance is created differently: to show the impact of correlation between the costs and the risks, we created the instance uncorrelated\_4 with the same graph and risks as for the instance lazio\_4\_1 but where the costs of each arc are the reciprocal of the risk of the first commodity. Figure~\ref{fig:correlation} shows those differences between the costs and risks for the instances lazio\_4\_1 and uncorrelated\_4.

@@

\begin{figure}
	\centering
	\includegraphics[width=0.48\columnwidth]{figs/correlationLazio41.eps}
	\hfill
	\includegraphics[width=0.48\columnwidth]{figs/correlationUncorrelated4.eps}
	\caption{\label{fig:correlation} Correlation between the costs and the risks of the first commodity for both the lazio\_4\_1 instance (left) and the uncorrelated\_4 instance (right). A single point corresponds to a single arc.}
\end{figure}

@@

\subsection{Convergence Towards the Pareto Front}
Figure~\ref{fig:hypervolumeOverTime} shows the boxplots of the hypervolume values for HypE with population size 50 and 100 for different number of generations. What can be seen is that, in general, the hypervolume indicator is mazimized over time and that for the algorithm with population size 50, the hypervolume value starts to stagnate starting from around generation 1500 with the mean still going up but the variance staying high. The fact that all runs found a similar hypervolume value which is significantly higher than the one of the random population in the beginning (around $9.62\cdot 10^26$ for both algorithms) indicates that the algorithm found solution sets which are close to the Pareto front.

\begin{figure}
	\centering
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_50.eps}
	\hfill
	\includegraphics[width=0.48\columnwidth]{../experiments/randVsCost/hypervolumes/hypervolumeOverTime_100.eps}
	\caption{\label{fig:hypervolumeOverTime} Boxplots of hypervolume values over time for numbers of generations in $[250, \ldots, 2000]$ for HypE with population size 50 (left) and population size 100 (right) on the lazio\_4 instance.}
\end{figure}

\subsection{Random Initialization Versus Cost-Optimal Initialization}
In order to see the impact of the two different initialization schemes \emph{random} and \emph{cost-optimal}, we performed 10 runs of HypE with both initialization schemes for 2,000 generations and a population size of 50 as above. Figure~\ref{fig:randVsCostOriginal} shows the resulting non-dominated solutions of the final populations. Surprisingly, the cost-optimal initialization could not produce other solutions than the cost-optimal one itself. As we will see below, the reason is a high correlation between the risk and costs of each arc, which seemingly result in the fact that no non-dominated solution can be reached from the initial solution by the applied mutation operator. However, the comparison between the both initialization schemes on the lazio\_4\_1 instance shows that also the algorithm employing the random initialization can find solutions which are close to the Pareto front---improving over the hypervolume of the single cost-optimal solution in later generations, see the left-hand plot of Fig.~\ref{fig:costVsRandBoxPlots}.

To show that the cost-optimal initialization is not useless, we run the algorithms again on the instance uncorrelated\_4 where the cost and risk values are fully uncorrelated. In this case, there are more local changes of the truck paths that result in non-dominated solutions and eventually in a larger variaty of solutions found by the algorithm employing the cost-optimal initialization (see Fig.~\ref{fig:randVsCostNew}). Due to the fact that the cost-optimal initialization starts already on the Pareto-front, it also results in higher hypervolume values than for the random initialization variant---both for a low and a high number of generations, see the right-hand side plot of Fig.~\ref{fig:costVsRandBoxPlots}.

@@

\subsection{Changing the Population Size}
In the next experiment, we aim at showing that changing the population size of the algorithm in standard ranges does not change the algorithm outcome drastically. To this end, Fig.~\ref{fig:popsizes} shows the nondominated solutions of 10 independent runs after 100,000 function evaluations for three different versions of HypE where the population size changes from 50 over 100 to 200 on the largest instance lazio\_4\_1 with 4 commodities. From the visual inspection of the resulting solutions, we cannot make a statement about which population size to favour. When comparing the hypervolume values of the resulting populations, however, an effect is visible. Figure~\ref{fig:algoComparison} shows the corresponding box plots of the hypervolume values and one-sided Wilcoxon rank sum tests on the populations' hypervolume indicator values reveil statistical significant differences between two of the three possible algorithm pairs. With a $p$-value of $0.02163$, the null hypothesis of no differences between a population size of 50 and a population size of 100 has to be rejected in favor of stating that the algorithm variant with population size 50 is superior to the one with population size 100. For the comparison between population size 50 and population size 200, the $p$-value equals $0.0144$ with the population size 50 being better whereas the comparison between population size 100 and population size 200 does not reveal a statistical difference (the $p$-values of the one-sided Wilcoxon tests are $0.2179$ and $0.8035$ respectively).

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{S_FW}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The transportation of hazmats is an important optimization problem in the field of sustainable development and in particular the equitable distribution of risks is of high interest. Within this study, we formalize this transportation problem as the minimization of three objectives and propose to use an evolutionary multiobjective optimization (EMO) algorithm to cope with the non-linear equity risk objective.

The algorithm, based on the state-of-the-art environmental selection scheme of HypE, employs a problem-specific variable-length representation and a mutation operator which are known to be well-suited for shortest path problems from theoretical studies. With its random initialization, the algorithm allows to find sets of non-dominated solutions in reasonable time which are close to the Pareto-optimal front. For the tested real-world instances with the road network of the Lazio region and dozens of trucks for up to 4 hazmat commodities, the current Java implementation allows to finish the 100,000 function evaluations of our experiments within less than 15 minutes on a standard Intel Core i5 laptop.

In future work, one might want to look further into adding (simple) heuristics within the EMO algorithm in order to improve the convergence of the algorithm. For example, one can observe that the solutions produced by the algorithm sometimes uses one and the same arc within a truck path, hence introducing cylces. With the three objectives cost, total and equity risk, such cycles can be easily prevented within the variation operator which might eventually result in a faster convergence of the algorithm to the Pareto front.
